---
title: "Practical: Other useful models"
subtitle: "Physalia workshop on GLMs"
author: "Bert van der Veen"
output: html_document
editor_options: 
  chunk_output_type: console
---

# R-packages

Whatever you can think of!

# Description

This is the final practical of the workshop. Here, you can combine everything you have learned so far into the analysis of a single dataset. You will pick a dataset, and analyse it from a-z:

1) data exploration
2) decide on a research question or hypothesis 
3) choose a response distribution
4) fit a model
5) perform a model comparison
6) find a valid model

You can choose from the following datasets:

1. lime: foliage measurements of small-leaved lime trees
2. brachy: cover class data for Brachythecium rutabulum with multiple covariates
3. r6data: data of two plant species under low and high nitrate supply. Response is the proportion of biomass in roots ("LMF"), stems ("SMF"), and leaves ("LMF").
4. andrew: Effect grazing sea urchin (density) on the colonization of filamentous algae (cover in 0.25m2 plots subject to sea urchin removal treatment).
5. any of the count data covered in the practicals
6. mistletoe: count of mistletoe infections in trees

Each of the datasets has unique properties. You might need to use a modeling approach in a package that we have not used so far. *Hint: the glmmTMB R-package has distributions appropriate for the properties or -most- of these data (not the brachy data, that is even more special).*. Don't forget to use `set.seed` for reproducibility if you do simulations. You can answer the questions below for guidance. You can construct the plots from the presentation yourself with the aforementioned functions, or use the `plot` function on your model.

# Tasks

1. Create some plots of the data. Are there any properties that stand out?

```{r}
lime <-read.csv("/home/bertv/GLM-workshop/data/lime.csv")
plot(Foliage ~ DBH, data = lime)

xi <- tweedie::tweedie.profile(Foliage~DBH, xi.vec = seq(from = 1.01, 3.5, length.out = 100), data = lime)
plot(xi)
abline(h = xi$L.max, v = xi$xi.max, col = "red")
# select something close to xi$xi.max to generate starting values for the glm algorithm
model <- glm(Foliage~DBH, family = statmod::tweedie(var.power = 1.7),data = lime)
# need to provide starting values, doesnt start otherwise
model1 <- glm(Foliage~DBH, family = statmod::tweedie(var.power = xi$xi.max), mustart = predict(model,type="response"), data = lime)
model2 <- glmmTMB::glmmTMB(Foliage~DBH, family = glmmTMB::tweedie, data = lime) # Power par = 1.9
```

2. What is your research question or hypothesis?

3. What range does the response variable take?

4. Which response distribution will you analyse the data with?

5. Fit one or multiple models and decide on the "final" model.

6. Perform model validation on the model of your choice.

7. Write up a summary of the results from the model.

8. Pick a different dataset and repeat!


```{r}
mistletoe <- read.csv("/home/bertv/GLM-workshop/data/MistletoeV2.csv")
plot(Infected~DBH, data=mistletoe)
plot(Infected~Distance.from.infected.stem, data=mistletoe)
plot(Infected~Min.distance.to.mistletoe.50cm.in.diameter, data=mistletoe)

model <- glm(Infected~DBH+Min.distance.to.mistletoe.50cm.in.diameter, family = "binomial", data = mistletoe)
plot(DHARMa::simulateResiduals(model))
model1 <- glm(Infected~Min.distance.to.mistletoe.50cm.in.diameter+offset(log(N12)), family = "binomial", data = mistletoe)
AIC(model, model1)
plot(ggeffects::predict_response(model, ~Min.distance.to.mistletoe.50cm.in.diameter+DBH))+ggplot2::theme_classic()
```

