---
title: "Practical: Generalised Linear Mixed Models"
subtitle: "Physalia workshop on GLMs"
author: "Bert van der Veen"
output: html_document
---

# Background

So far, all models we have fitted assume that observations are independent. In practice, data often has a grouped or hierarchical structure: repeated measurements on the same individual, observations from the same plot or nest, or samples from the same site. Ignoring this structure inflates our effective sample size and leads to overconfident inference.

Generalised Linear Mixed Models (GLMMs) extend GLMs by including **random effects** to account for this non-independence. In this practical we will use the `Owls` dataset from the `glmmTMB` package, which records the number of sibling negotiation calls made by chicks in owl nests, along with information on food treatment, brood size, arrival time of the parent, and sex of the parent.

```{r, message=FALSE, warning=FALSE}
library(glmmTMB)
data(Owls)
str(Owls)
head(Owls)
```

The key grouping variable is `Nest`: chicks from the same nest share an environment and genetics, so their observations are not independent.

# Distribution of the response

```{r, fig.width=10}
hist(Owls$SiblingNegotiation, breaks=30,
     xlab="Sibling negotiation (count)", main="Distribution of response")
```

The distribution is right-skewed and count-valued. Compare the mean and variance:

```{r}
mean(Owls$SiblingNegotiation)
var(Owls$SiblingNegotiation)
```

The variance is much larger than the mean, indicating overdispersion relative to a Poisson distribution.

# Variation across nests

```{r, fig.width=10}
boxplot(SiblingNegotiation ~ Nest, data=Owls,
        xlab="Nest", ylab="Sibling negotiation",
        las=2, cex.axis=0.7, main="Response by nest")
```

There is considerable variation between nests. A model that ignores this grouping will treat all observations as independent, inflating the effective sample size.

# A naive Poisson GLM

Let us start by fitting a Poisson GLM that ignores the nest grouping. We include `FoodTreatment` and `ArrivalTime` as fixed effects.

```{r}
m_pois <- glm(SiblingNegotiation ~ FoodTreatment + ArrivalTime,
              family = poisson, data = Owls)
summary(m_pois)
```

We can check for overdispersion using the residual deviance divided by residual degrees of freedom. Values much larger than 1 indicate overdispersion:

```{r}
deviance(m_pois) / df.residual(m_pois)
```

```{r, message=FALSE, warning=FALSE, fig.width=10}
library(DHARMa)
sim_pois <- simulateResiduals(m_pois, plot=TRUE)
```

```{r, fig.width=6}
testDispersion(sim_pois)
```

The residuals show strong patterns and the dispersion test confirms overdispersion, consistent with ignoring both the nest structure and the count distribution.

# A negative-binomial GLMM

We now account for both overdispersion (using a negative-binomial distribution) and the nest structure (using a random intercept for `Nest`).

```{r}
m_nb <- glmmTMB(SiblingNegotiation ~ FoodTreatment + ArrivalTime + (1|Nest),
                family = nbinom2, data = Owls)
summary(m_nb)
```

The `(1|Nest)` term adds a nest-specific random intercept: each nest gets its own baseline level of calling, drawn from a normal distribution with mean 0 and an estimated variance. A larger variance implies more between-nest differences.

## Extracting random effects

```{r}
ranef(m_nb)
```

```{r, fig.width=10, message=FALSE, warning=FALSE}
library(ggeffects)
plot(predict_response(m_nb, terms = "Nest", type = "random",
                      interval = "confidence")) +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, vjust = 1, hjust = 1))
```

Which nests have higher than average calling, and which lower?

# Residual checking with DHARMa

For GLMMs, standard residuals are not appropriate. The `DHARMa` package provides simulation-based quantile residuals that work for any GLMM.

```{r, message=FALSE, warning=FALSE, fig.width=10}
sim_res <- simulateResiduals(m_nb, plot=TRUE)
```

# Comparing models

Let us check whether the random intercept for `Nest` improves the model by comparing to a negative-binomial GLM without random effects.

```{r}
m_nb_fixed <- glmmTMB(SiblingNegotiation ~ FoodTreatment + ArrivalTime,
                      family = nbinom2, data = Owls)
anova(m_nb_fixed, m_nb)
```

# Adding brood size

The number of chicks in a nest (`BroodSize`) is a natural offset: more chicks means more calls regardless of food treatment. We can include it as a log-offset.

```{r}
m_nb2 <- glmmTMB(SiblingNegotiation ~ FoodTreatment + ArrivalTime + (1|Nest),
                 offset = log(BroodSize),
                 family = nbinom2, data = Owls)
summary(m_nb2)
```

How do the fixed effect estimates change when you include the brood size offset?

# Exercises

<details>
<summary>Click to see the exercises</summary>

1. In the naive Poisson GLM, what is the overdispersion factor (residual deviance / df)? Is it consistent with what you see in the histogram of counts?

2. Fit a Poisson GLMM (keeping the random intercept but using `family = poisson`). How do its fixed effect estimates compare to the negative-binomial GLMM? Check the DHARMa residuals: is there remaining overdispersion?

3. An alternative to the negative-binomial for handling overdispersion in a Poisson GLMM is to add an **observation-level random effect** (OLRE): a random intercept with one level per observation. Add a unique row ID to the data and fit `SiblingNegotiation ~ FoodTreatment + ArrivalTime + (1|Nest) + (1|obs)` with `family = poisson`. Does this resolve the overdispersion? How does it compare to the negative-binomial GLMM?

4. Add `SexParent` as a fixed effect to the negative-binomial GLMM with the offset. Is it an important predictor?

5. Try fitting a model with both random intercepts and random slopes for `ArrivalTime` by `Nest` using `(ArrivalTime|Nest)`. Does it converge? Does the additional complexity seem warranted based on the summary output?

6. Use `ggeffects::predict_response()` or `emmeans` to visualise the predicted effect of `FoodTreatment` on the expected count (marginalising over nests). Do food-deprived nests call more than satiated ones?

7. What would happen if you ignored the nest random effect entirely? Fit a standard NB GLM and compare the standard errors of the fixed effects to those from the GLMM. Are they larger or smaller in the GLM?

</details>
