---
title: "Practical: Generalized Additive Models"
subtitle: "Physalia workshop on GLMs"
author: "Bert van der Veen"
output: html_document
---

# Background

GLMs assume that covariates enter the linear predictor **linearly**: a one-unit change in $x$ always produces the same change in $g\{\text{E}(y)\}$, regardless of where on the $x$-axis you are. Many ecological relationships are non-linear, and a linear model misspecifies the functional form, leading to systematic residual patterns.

Generalized Additive Models (GAMs) replace the linear term $x\beta$ with a smooth function $f(x)$ estimated from the data. The two main packages for fitting GAMs in R are:

- `mgcv`: the standard GAM package by Simon Wood
- `gratia`: provides `ggplot2`-based visualisation tools for `mgcv` models

In this practical we will use the ISIT dataset (bioluminescent sources at depth in the ocean) from Gillibrand et al. (2007), the same dataset used in the lecture.

```{r, message=FALSE, warning=FALSE}
library(mgcv)
library(gratia)

ISIT <- read.csv("../../data/ISIT.csv")
str(ISIT)
```

The response is `Sources` (count of bioluminescent organisms), and the main predictor is `SampleDepth` (depth in metres). There are 19 stations; for now we will focus on a single station.

```{r}
dat <- subset(ISIT, Station == 9)
nrow(dat)
```

# Distribution of the response

```{r, fig.width=10}
hist(dat$Sources, breaks=20,
     xlab="Bioluminescent sources", main="Distribution of response")
```

```{r, fig.width=10}
plot(Sources ~ SampleDepth, data=dat, pch=16, col="steelblue",
     xlab="Depth (m)", ylab="Bioluminescent sources",
     main="Station 9: sources vs. depth")
```

Does the relationship look linear? What distribution might be appropriate for the response?

# A GLM baseline

Let us first fit a Gamma GLM with a log link, treating depth as a linear predictor.

```{r}
m_glm <- glm(Sources ~ SampleDepth, family=Gamma(link="log"), data=dat)
summary(m_glm)
```

```{r, fig.width=10}
xseq <- seq(min(dat$SampleDepth), max(dat$SampleDepth), length.out=300)
pred_glm <- predict(m_glm, newdata=data.frame(SampleDepth=xseq), type="response")

plot(Sources ~ SampleDepth, data=dat, pch=16, col="gray60",
     xlab="Depth (m)", ylab="Sources", main="GLM fit")
lines(xseq, pred_glm, col="red", lwd=2)
```

```{r, fig.width=10}
plot(log(fitted(m_glm)), residuals(m_glm, type="deviance"),
     xlab="log(Fitted values)", ylab="Deviance residuals",
     main="GLM residuals")
abline(h=0, lty="dashed", col="red")
lines(loess.smooth(log(fitted(m_glm)), residuals(m_glm, type="deviance")),
      col="blue", lwd=2)
```

The pattern in the residuals suggests the linear predictor does not capture the shape of the relationship.

# Fitting a GAM

We now fit a GAM with a smooth term for `SampleDepth`. We use `bs="cr"` (cubic regression spline) and `method="REML"` for smoothing parameter selection.

```{r}
m_gam <- gam(Sources ~ s(SampleDepth, bs="cr"),
             family=Gamma(link="log"), data=dat, method="REML")
summary(m_gam)
```

The `edf` column in the smooth terms table shows the **effective degrees of freedom**: a measure of how wiggly the estimated smooth is. An edf of 1 is a straight line; higher values indicate more curvature.

## Plotting the smooth with gratia

The `gratia` package provides `draw()` for visualising GAM smooths with confidence intervals.

```{r, fig.width=8}
draw(m_gam)
```

```{r, fig.width=8}
draw(m_gam, residuals=TRUE)
```

# Model checking

```{r, fig.width=10}
appraise(m_gam, method="simulate")
```

# Comparing GLM and GAM

```{r, fig.width=10}
pred_gam <- predict(m_gam, newdata=data.frame(SampleDepth=xseq), type="response")

plot(Sources ~ SampleDepth, data=dat, pch=16, col="gray60",
     xlab="Depth (m)", ylab="Sources", main="GLM vs. GAM")
lines(xseq, pred_glm, col="red", lwd=2, lty="dashed")
lines(xseq, pred_gam, col="orange", lwd=3)
legend("topright", lty=c("dashed","solid"), col=c("red","orange"),
       legend=c("GLM", "GAM"), lwd=2)
```

```{r}
AIC(m_glm, m_gam)
```

# Extending to all stations

The ISIT data has 19 stations, and observations from the same station are not independent. We can extend the GAM to include a random intercept for station, giving a GAMM (Generalized Additive Mixed Model).

```{r}
# Full dataset includes zeros so we use Tweedie
m_gamm <- gam(Sources ~ s(SampleDepth, bs="cr") + s(Station, bs="re"),
              family=tw(), data=ISIT, method="REML")
summary(m_gamm)
```

The `s(Station, bs="re")` term is a random intercept expressed as a smooth, as introduced in the lecture.

```{r, fig.width=10}
draw(m_gamm, select=1)
```

How does the depth smooth change when you include all stations compared to fitting only Station 9?

# Exercises

<details>
<summary>Click to see the exercises</summary>

1. Refit the GAM for Station 9 with a thin plate spline (`bs="tp"`, the default) instead of `bs="cr"`. Does the fitted smooth look different? Compare AIC.

2. Fit a GAM to Station 9 using a Poisson family instead of Gamma. Check the diagnostics with `gratia::appraise()`. Does a Poisson distribution seem appropriate?

3. For the full ISIT dataset, fit a model with a station-specific smooth for depth: `s(SampleDepth, by=factor(Station), bs="cr", k=5)`. Note that some stations have very few observations (Station 4 has only 5), so a small `k` is needed to avoid fitting problems. How does this compare (AIC) to the model with a global smooth plus random intercept?

4. Use `gratia::appraise()` on your best model. Are there any remaining problems?

5. Try the same GAMM in `glmmTMB`:
```r
library(glmmTMB)
m_tmb <- glmmTMB(Sources ~ s(SampleDepth, bs="cr") + (1|Station),
                 family=Gamma(link="log"), data=ISIT)
```
Compare the predicted smooth from `glmmTMB` and `mgcv`. Are they similar?

6. The dataset has a `Season` column. Add `Season` as a factor to the GAMM. Does the seasonal effect improve the model?

</details>
