---
title: "Sampling and variation"
subtitle: "Physalia workshop on GLMs"
author: "Bert van der Veen"
output: html_document
---

# Background

In the lecture we used an orchid example to introduce sampling variation and maximum likelihood estimation. We imagined walking through a field and recording whether each spot had an orchid or not, binary data that follows a binomial distribution.

The binomial distribution has two parameters: `size` (the number of picks per field visit, $N$) and `prob` (the true probability of finding an orchid, $p$). In R, `rbinom(n, size, prob)` simulates `n` field visits, each with `size` picks.

```{r}
y <- rbinom(n = 20, size = 10, prob = 0.4)
y
```

Each value is the number of orchids found in 10 picks. Running this again gives different results, that is sampling variation.

The likelihood tells us how probable our observed data is for a given value of $p$. `dbinom` computes it directly, we just need to sum over all observations and maximise over $p$. We can do this with `optimize`, just as in the lecture:

```{r}
mle <- optimize(function(p) sum(dbinom(y, size = 10, prob = p, log = TRUE)),
                lower = 0, upper = 1, maximum = TRUE)
mle$maximum
```

# Part 1

Here is what I want you to do:

1. Secretly choose values for `prob` (between 0 and 1), `size` (picks per visit), and `n` (number of visits). Do not tell the other groups.
2. Simulate data using `rbinom` and plot a histogram with `hist`.
3. Find the MLE using `optimize` and `dbinom`, as shown in the background above, just swap in your own `size` and `y`.
4. Repeat the simulation many times and look at the distribution of the MLE. Simulate `B = 500` datasets using your chosen parameters, find the MLE for each, and plot a histogram:

```{r, eval = FALSE}
B <- 500
mles <- numeric(B)
for (i in 1:B) {
  ysim <- rbinom(n = 20, size = 10, prob = 0.4)  # use your own n, size, prob
  fit <- optimize(function(p) sum(dbinom(ysim, size = 10, prob = p, log = TRUE)),
                  lower = 0, upper = 1, maximum = TRUE)
  mles[i] <- fit$maximum
}
hist(mles, xlab = "MLE of p", main = "Sampling distribution of the MLE")
abline(v = 0.4, col = "red")  # add your true prob here
```

What shape does the distribution take? What happens if you increase `n` or `size`? How does this connect to what was shown in the lecture about the Central Limit Theorem?

5. Save your simulated data and share the file via Slack, do not reveal your parameter values:

```{r, eval = FALSE}
save(y, file = "y.RData")
```

6. Show your histogram on Zoom. We will discuss as a group before opening Part 2.

# Part 2

<details><summary> Open after discussion on zoom</summary>

You have received a file from another group. Load it with:

```{r, eval = FALSE}
load("y.RData")
```

Make sure the file is in your working directory, you can check with `getwd()`.

Now find the MLE for $p$ from their data. The `size` parameter can be read off from the data, what is `max(y)`?

```{r, eval = FALSE}
size <- max(y)   # or eyeball it from the histogram
mle <- optimize(function(p) sum(dbinom(y, size = size, prob = p, log = TRUE)),
                lower = 0, upper = 1, maximum = TRUE)
mle$maximum
```

You can also compute two confidence intervals, just as the lecture showed. The CLT slide gives an asymptotic normal CI:

```{r, eval = FALSE}
n_total <- length(y) * size       # total picks: n_times * n_picks
se      <- sqrt(mle$maximum * (1 - mle$maximum) / n_total)
ci_normal <- c(mle$maximum - 1.96 * se, mle$maximum + 1.96 * se)
ci_normal
```

Because we also know the exact distribution — $\sum y_i \sim \text{Binom}(n_{total},\, p)$ — we can use it directly for a CI that does not rely on the normal approximation:

```{r, eval = FALSE}
ci_exact <- c(qbinom(0.025, n_total, mle$maximum) / n_total,
              qbinom(0.975, n_total, mle$maximum) / n_total)
ci_exact
```

Compare the two. Do they agree? When might they differ?

Discuss with your group:

1. Does your confidence interval seem narrow or wide? What determined that?
2. Which group do you think simulated the data? Think back to the histograms on Zoom.
3. After the groups reveal their true parameters: did groups that used larger `n` or `size` get identified more easily or less easily? What does this tell you about the relationship between sample size and the precision of a parameter estimate?

</details>

[If you finish early, there is also the 2024 edition of this practical.](https://htmlpreview.github.io/?https://github.com/BertvanderVeen/GLM-workshop/blob/main/Practicals/2024/1Practical.html)
