---
title: "Sampling and variation"
subtitle: "Physalia workshop on GLMs"
author: "Bert van der Veen"
output: html_document
---

# Background

In the lecture we used an orchid example to introduce sampling variation and maximum likelihood estimation. We imagined walking through a field and recording whether each spot had an orchid or not, such binary data that follows a binomial distribution.

The binomial distribution has two parameters: `size` (the number of picks per field, $n_{picks}$) and `prob` (the true probability of finding an orchid, $p$). In R, `rbinom(n.fields, size = n.picks, prob)` simulates `n.fields` field visits, each with `n.picks` picks.

Everyone uses `n.picks = 10` picks per field.

```{r}
n.picks <- 10
y <- rbinom(n = 1, size = n.picks, prob = 0.4)
y
```

The value is the number of orchids found in 10 picks. Running this again gives different results, that is sampling variation.

The likelihood tells us how probable our observed data is for a given value of $p$. `dbinom` computes it directly; we sum over all observations and maximise over $p$ using `optimize`:

```{r}
mle <- optimize(function(p) sum(dbinom(y, size = n.picks, prob = p, log = TRUE)),
                lower = 0, upper = 1, maximum = TRUE)
mle$maximum
```

With more data the MLE gets closer to the truth:

```{r}
y_more <- rbinom(n = 200, size = n.picks, prob = 0.4)
mle_more <- optimize(function(p) sum(dbinom(y_more, size = n.picks, prob = p, log = TRUE)),
                     lower = 0, upper = 1, maximum = TRUE)
mle_more$maximum
```

Once we have the MLE we can quantify uncertainty with confidence intervals. The normal CI uses the asymptotic standard error from the CLT:

```{r}
n.fields  <- length(y)
n_total   <- n.fields * n.picks
se        <- sqrt(mle$maximum * (1 - mle$maximum) / n_total)
ci_normal <- c(mle$maximum - 1.96 * se, mle$maximum + 1.96 * se)
ci_normal
```

Because $\sum y_i \sim \text{Binom}(n_{total}, p)$ exactly, we can also use the exact binomial distribution directly:

```{r}
ci_exact <- c(qbinom(0.025, n_total, mle$maximum) / n_total,
              qbinom(0.975, n_total, mle$maximum) / n_total)
ci_exact
```


# Part 1

Here is what I want you to do:

1. Choose values for `prob` (between 0 and 1) and `n.fields` (number of fields to sample). Everyone uses `n.picks = 10`.
2. Simulate data using `rbinom` and plot a histogram with `hist`.
3. Find the MLE using `optimize` and `dbinom`. Then repeat with ten times as many fields. Does the MLE get closer to the truth?
4. Compute both confidence intervals for your MLE. Do the normal and exact CIs agree?
5. What happens to the intervals when you increase `n.fields`?
6. What do you expect to happen if `prob` were close to 0 or 1? Try it.
7. One person per group: save your simulated data and share the file via Slack. Do not reveal your parameter values to the other groups:

```{r, eval = FALSE}
save(y, file = "y.RData")
```

8. Show your histogram on Zoom. We will discuss as a group before opening Part 2.

# Part 2

<details><summary> Open after discussion on zoom</summary>

You have received a file from another group. Load it with:

```{r, eval = FALSE}
load("y.RData")
```

Make sure the file is in your working directory, you can check with `getwd()`.

1. Plot a histogram of the data. Can you guess the true `prob` and `n.fields` used by the other group?
2. Find the MLE and compute both confidence intervals as shown in the background.
3. What is your MLE estimate of `prob`? Interpret it in plain language: what does it mean for the orchid field?
4. Do the normal and exact CIs agree? What drives any difference between them?
5. After the groups reveal their true parameters: does your CI contain the true `prob`? Did groups with larger `n.fields` get narrower intervals, and were their parameters easier to identify?

</details>

[If you finish early, there is also the 2024 edition of this practical.](https://htmlpreview.github.io/?https://github.com/BertvanderVeen/GLM-workshop/blob/main/Practicals/2024/1Practical.html)
